---
title: "UMN Day of Data 2021"
output: html_notebook
---

# Install missing packages

If you're not sure whether you have the **ipumsr**, **knitr**, and **tidyverse**
packages installed, run the code chunk below:

```{r eval=FALSE}
if (!require(ipumsr)) install.packages("ipumsr")
if (!require(knitr)) install.packages("knitr")
if (!require(tidyverse)) install.packages("tidyverse")
```

# Load packages

This script uses functions from the **ipumsr** and **knitr** packages, as well as
the **tidyverse** collection of packages, so we'll attach those packages now.

```{r}
library(ipumsr)
library(knitr)
library(tidyverse)
```

# Read in the data

First, let's load our data, along with the metadata in the XML DDI file. Let's
set `lower_vars` to `TRUE` here to make all variable names lowercase, for ease
of typing.

```{r results='hide'}
metadata <- read_ipums_ddi("dod2021_ipums_acs_cities.xml", lower_vars = TRUE)
data <- read_ipums_micro(metadata, data_file = "dod2021_ipums_acs_cities.csv.gz")
```

# Explore the data

The code below prints out a list of all the variables in our dataset, along 
with a short description of each. You can refer to this list if you want to 
explore these data in your own way. (Note that we use the function `kable()` 
from package **knitr** here simply because it does a better job of fitting the 
whole table on the screen than does the default table output.)


```{r}
metadata %>% 
  ipums_var_info() %>% 
  select(var_name, var_label) %>% 
  arrange(var_name) %>% 
  kable()
```


If you want to see the values, value labels, and frequencies for any of these
variables, you can use some code like the following (`levels = "both"` creates 
values that include both the numeric code and the value label).

```{r}
data %>% 
  mutate(educ = as_factor(educ, levels = "both")) %>% 
  count(educ) %>% 
  arrange(educ)
```

The visualizations below will aggregate the data at the level of cities and 
regions, so let's take a look at the cities we have in our dataset, organized by 
region.

```{r}
data %>% 
  mutate(
    region_name = as_factor(region, levels = "both"),
    state_name = as_factor(statefip, levels = "both"), 
    city_name = as_factor(city, levels = "both")
  ) %>%
  distinct(region_name, state_name, city_name) %>%
  arrange(region_name, state_name, city_name) %>% 
  kable()
```

In the city-level visualizations below, we will focus on cities in the West 
North Central Division (including Minneapolis and St. Paul), but you should 
feel free to modify the code to explore any region or cities that you are 
interested in.

# Prepare the data

To simplify the visualizations below, it will be helpful to recode and collapse 
some variables into fewer categories. Let's do that here so that we can focus 
on the visualization code below.

First, we'll create `city_name` and `region_name` variables so it's easy to
print city and region names instead of codes. We can still use the original 
variables to refer to the codes for those variables.

```{r}
data <- data %>% 
  mutate(city_name = as_factor(city), region_name = as_factor(region))
```

Some of our visualizations below will examine mode of transportation to work, so 
let's create a simplified version of that variable.

```{r}
data <- data %>% 
  mutate(
    tranwork_collapsed = case_when(
      tranwork == 0 ~ NA_character_,
      tranwork == 10 ~ "Automobile",
      tranwork %in% 31:39 ~ "Public transit",
      tranwork == 50 ~ "Bicycle",
      tranwork == 60 ~ "Walking",
      tranwork %in% c(20, 70) ~ "Other",
      tranwork == 80 ~ "Worked at home"
    )
  )
```

Finally, let's create a collapsed version of the race variable.

```{r}
data <- data %>% 
  mutate(
    race_collapsed = case_when(
      race == 1 ~ "White",
      race == 2 ~ "Black",
      race == 3 ~ "American Indian or Alaska Native",
      race %in% 4:6 ~ "Asian or Pacific Islander",
      race %in% 7:9 ~ "Other or multiple races"
    )
  )
```


# Visualize the data

## Number of working age people

As a first pass at the data, let's look at how many people of prime working age 
(25-54) are in our data for each year. (Note that this includes both employed 
and unemployed persons; to limit to employed persons, you could use the 
variable "empstat".)

```{r}
data %>% 
  filter(age %in% 25:54) %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) + 
    geom_line(size = 2) + 
    geom_point(size = 3) +
    ggtitle("Number of people ages 25-54, 2013-2019")
```

Our first reaction to this plot might be to wonder why there are so many more
people in our data in the middle years of the time range. However, when we look
at the y-axis values, we see that the number of cases really only ranges from
about 104,000 to about 106,000. A bar plot, which always starts the y-axis at
zero, does a better job here of showing the (small) magnitude of variation over
time.

```{r}
data %>% 
  filter(age %in% 25:54) %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) + 
    geom_col() +
    ggtitle("Number of people ages 25-54, 2013-2019")
```

## Number of working age people by city

Before we dive into cities in a particular region, let's look at the range of 
number of workers across all cities in our data.

```{r}
data %>% 
  filter(age %in% 25:54) %>% 
  count(city_name) %>% 
  mutate(city_name = fct_reorder(city_name, n, mean)) %>% 
  ggplot(aes(x = city_name, y = n)) + 
    geom_col() + 
    coord_flip() + 
    labs(title = "Number of persons of working age in the data", x = NULL, 
         y = NULL)
```

This graph shows that a large portion of our dataset is made up of people from 
the few most populous cities.

## Number of working age people by city over time

Now focusing on the West North Central Division, let's see how many people of
working age are in our data for each city between 2013 and 2019.

```{r}
data %>% 
  filter(region == 22 & age %in% 25:54) %>% 
  count(city_name, year) %>% 
  mutate(city_name = fct_reorder(city_name, n, .fun = mean)) %>% 
  ggplot(aes(x = year, y = n, color = city_name)) + 
    geom_line(size = 2) + 
    geom_point(size = 3) + 
    guides(color = guide_legend(reverse = TRUE)) + # reorder legend labels
    labs(x = "Year", y = NULL, color = "City", 
         title = "Number of people ages 25-54")
```

The trends are pretty flat here, as we would expect, but we should notice that 
we have relatively few observations in each year from each city, especially 
Sioux Falls and Kansas City, so we should be cautious about breaking down those 
city-years further into groups by other things such as race or age.

## Modes of transportation across all cities

How do people across all these cities get to work? Keep in mind that these 
results will be strongly influenced by a few large cities.

```{r}
data %>% 
  filter(age %in% 25:54) %>% 
  filter(!is.na(tranwork_collapsed)) %>% 
  ggplot(aes(x = tranwork_collapsed)) +
    geom_bar() +
    labs(
      title = "Number of people using each mode of transportation to get to work", 
      x = NULL, 
      y = NULL
    )
```

Automobile transportation is clearly dominant, and while a large minority of 
people use public transit, this is heavily dependent on region and city. Besides 
automobile and public transit, walking and working from home are the next most 
common ways of getting to work, with bicycle and other modes relatively 
uncommon.

## Modes of transportation in Minneapolis

If we're interested in one particular city, we can filter down to that city and 
graph the number of people using each mode.

```{r}
data %>% 
  filter(age %in% 25:54) %>% 
  filter(city_name == "Minneapolis, MN") %>% 
  filter(!is.na(tranwork_collapsed)) %>% 
  ggplot(aes(x = tranwork_collapsed)) +
    geom_bar() +
    labs(
      title = paste0(
        "Number of people using each mode of transportation to get to work, ", 
        "Minneapolis"
      ),
      x = NULL, 
      y = NULL
    )
```


## Modes of transportation by region

Does the breakdown of transportation modes look different by region? To create 
a visualization by mode of transportation *and* region, it will be helpful to 
switch from visualizing the *number* of people using each mode to visualizing 
the *share* or *percentage* of people using each mode by region, since these 
regions have widely varying numbers of people in our dataset. 

The code below creates a summary dataset capturing the percentage of people in
each region using each mode of transportation across all years 2013-2019. Feel
free to gloss over this code if you want to focus on the visualization code, but
keep in mind that creating these types of summaries is often a necessary first
step to making visualizations from microdata.

```{r}
mode_share_by_region <- data %>% 
  filter(age %in% 25:54) %>% 
  filter(!is.na(tranwork_collapsed)) %>% 
  group_by(region_name) %>% 
  mutate(total_workers = n()) %>% 
  group_by(region_name, tranwork_collapsed) %>% 
  summarize(pct_mode_share = 100 * n() / total_workers[1], .groups = "drop")
```

Now we can graph the transportation mode share by region. Note that New Orleans,
LA is the only city in our data from the West South Central region.

```{r}
mode_share_by_region %>% 
  ggplot(aes(x = tranwork_collapsed, y = pct_mode_share)) + 
    geom_col() + 
    coord_flip() +
    facet_wrap(~region_name) + 
    labs(title = "Percentage of people using each transportation mode by region", 
         x = NULL, y = "%")
```

This graph reveals the wide variation of public transit usage by region, with 
public transit as the most common mode in the Middle Atlantic Division (which 
includes New York City), and fairly common in the New England, East North 
Central, and South Atlantic divisions, but relatively uncommon in the other 
regions.

## West North Central Division

### Modes of transportation by city 

Since we have relatively few cases in each year for each city, we should 
probably pool across years to look at differences in mode of transportation to 
work. Before creating a visualization, let's create a summary dataset that will 
serve as the basis for the graph.

Again, you can skim over this part of the code if you want to get to the 
visualization, but here's what the summary dataset looks like if you're 
interested.

```{r}
mode_share_by_city <- data %>% 
  filter(region == 22 & age %in% 25:54) %>% 
  filter(!is.na(tranwork_collapsed)) %>% 
  group_by(city_name) %>% 
  mutate(total_workers = n()) %>% 
  group_by(city_name, tranwork_collapsed) %>% 
  summarize(pct_mode_share = 100 * n() / total_workers[1], .groups = "drop")

mode_share_by_city
```

Now we can focus on the code to create the graph, below.

```{r}
mode_share_by_city %>% 
  ggplot(aes(x = tranwork_collapsed, y = pct_mode_share)) + 
    geom_col() + 
    coord_flip() +
    facet_wrap(~city_name) +
    labs(title = "Percentage of people using each transportation mode by city", 
         x = NULL, y = "%")
```

Automobiles are overwhelmingly the most common mode across all cities, so let's 
also make a graph that excludes automobile drivers to get a better picture of 
the relative popularity of other modes.

```{r}
mode_share_by_city %>% 
  filter(tranwork_collapsed != "Automobile") %>% 
  ggplot(aes(x = tranwork_collapsed, y = pct_mode_share)) + 
    geom_col() + 
    coord_flip() +
    facet_wrap(~city_name) +
    labs(title = "Percentage of people using each transportation mode by city", 
         x = NULL, y = "%")
```

### Public transit use by city and race

First let's create a summary dataset.

```{r}
mode_share_by_race <- data %>% 
  filter(region == 22 & age %in% 25:54) %>% 
  filter(!is.na(tranwork_collapsed)) %>% 
  group_by(city_name, race_collapsed) %>% 
  mutate(total_workers = n()) %>% 
  group_by(city_name, tranwork_collapsed, race_collapsed) %>% 
  summarize(pct_mode_share = 100 * n() / total_workers[1], .groups = "drop")
```

Now let's visualize public transit usage by city and race.

```{r}
mode_share_by_race %>% 
  filter(tranwork_collapsed == "Public transit") %>% 
  mutate(
    # reorder categories by public transit usage
    race_collapsed = fct_reorder(race_collapsed, pct_mode_share, .fun = mean),
    city_name = fct_reorder(city_name, pct_mode_share, .fun = mean)
  ) %>% 
  ggplot(aes(x = city_name, y = pct_mode_share, fill = race_collapsed)) +
    geom_col(position = position_dodge()) + # grouped bars side-by-side instead of stacked
    coord_flip() + # horizontal instead of vertical bars
    guides(fill = guide_legend(reverse = TRUE)) + # reorder legend labels
    labs(fill = "Race") + 
    xlab(NULL) +
    ylab("% mode share") + 
    ggtitle("Public transit usage by race in select cities, 2013-2019")
```

### Working at home by city and race

Now let's visualize rates of working at home by city and race.

```{r}
mode_share_by_race %>% 
  filter(tranwork_collapsed == "Worked at home") %>% 
  # Turns out there are only 13 persons with race of "American Indian or 
  #   Alaska Native" from Saint Louis in the data, so let's exclude that group
  filter(
    !(city_name == "Saint Louis, MO" & 
        race_collapsed == "American Indian or Alaska Native")
  ) %>% 
  mutate(
    # reorder categories by rates of working at home
    race_collapsed = fct_reorder(race_collapsed, pct_mode_share, .fun = mean),
    city_name = fct_reorder(city_name, pct_mode_share, .fun = mean)
  ) %>% 
  ggplot(aes(x = city_name, y = pct_mode_share, fill = race_collapsed)) +
    geom_col(position = position_dodge()) + # grouped bars side-by-side instead of stacked
    coord_flip() + # horizontal instead of vertical bars
    guides(fill = guide_legend(reverse = TRUE)) + # reorder legend labels
    labs(fill = "Race") + 
    xlab(NULL) +
    ylab("% of workers") + 
    ggtitle("Prevalence of working at home by race in select cities, 2013-2019")
```